### 2020-04-27 Case Studies/02_race-times.rmd
finished cherry blossom
finished cherry blossom

@@

@@

@@

>>>>

```{r}
groups <- 1 + (1:length(men8L) %% 9)

plot(x = 40, y = 60, type = "n",
     xlim = c(20, 80), ylim = c(40, 160),
     xlab = "Age (years)", ylab = "Run Time (minutes)")

addRunners <- function(listRunners, colors, numLty) {
   numRunners <- length(listRunners)

   colIndx <- 1 + (1:numRunners) %% length(colors)
   ltys <- rep(1:numLty, each = length(colors), length = numRunners)

   mapply(function(df, i) {
      lines(df$runTime ~ df$age,
            col = colors[colIndex[i]], lwd = 2, lty = ltys[i])
   }, listRunners, i = 1:numRunners)
}

colors <- c("#e41alc", "#377eb8", "#4daf4a", "#984ea3",
            "#ff7f00", "#a65628")

par(mfrow = c(3, 3), mar = c(2, 2, 1, 1))
invisible(
   sapply(1:9, function(grpId) {
      plot(x = 0, y = 0, type = "n",
           xlim = c(20, 80), ylim = c(50, 130),
           xlab = "Age (years)", ylab = "Run Time (minutes)")

      addRunners(men8L[ groups == grpId ], colors, numLty = 6)
   })
)
```

```{r}
fitOne <- function(oneRunner, addLine = F, col = "grey") {
   lmOne <- lm(runTime ~ age, data = oneRunner)

   if ( addLine )
      lines(x = oneRunner$age, y = predict(lmOne),
            col = col, lwd = 2, lty = 2)

   ind <- floor( (nrow(oneRunner) + 1) / 2)
   res <- c(coefficients(lmOne)[2], oneRunner$age[ind],
            predict(lmOne)[ ind ])

   names(res) <- c("ageCoeff", "medAge", "predRunTime")

   return(res)
}
```

```{r}
par(mfrow = c(1, 1))
plot( x = 0, y = 0, type = "n",
      xlim = c(20, 80), ylim = c(50, 130),
      xlab = "Age (years)", ylab = "Run Time (minutes)")

addRunners( men8L[ groups == 9 ], colors, numLty = 6)
lapply(men8L[groups == 9], fitOne, addLine = T, col = "black")
```

```{r}
men8LongFit <- lapply(men8L, fitOne)

coeffs <- sapply(men8LongFit, "[", "ageCoeff" )
ages <- sapply(men8LongFit, "[", "medAge")

longCoeffs <- lm(coeffs ~ ages)

summary(longCoeffs)
```

@@

```{r}
data_paths <- sapply(c("men_txt", "women_txt"), function(p) {
   path <- file.path(data.dir, p)

   sapply(path, function( f ) { 
      file.path(path, list.files(path) )
   })
})

consolidated_files <- c(data_paths[, 1], data_paths[, 2])

tp <- consolidated_files[1]

files <- lapply(consolidated_files, function(f) read.fwf(f, widths = 120))

length(files)
```

@@

sapply(womenDF, function(x) sum(is.na(x$runTime)))

@@

### 8.)

Modify the call to the _plot()_ function that created figure 2.6 to create Figure 2.7. To do this, read the documentation for plot() to determine which parameters could be helpful.

```{r}
ggplot(cbMen, aes(age, runTime)) +
   geom_jitter(col = Purple8A) +
   geom_smooth() +
   xlim(10, 75) +
   ylim(45, 150)
```

### 9.)

Modify the piecewise linear fit from Section 2.4.2 to include a hing a 70. Examine the coefficients from the fit and compare the fitted curve to the loess curve. Does the additional hing improve the fit?

```{r}
age20to70 <- seq( from = 20, to = 70, by = 1)

mR.lo992 <- loess(runTime ~ age, cbMenSub[ cbMenSub$year == 1999,])
mR.lo.pr992 <- predict(mR.lo99, data.frame(age = age20to70))

summary(mR.lo.pr992)

mR.lo122 <- loess(runTime ~ age, cbMenSub[ cbMenSub$year == 2012, ])
mR.lo.pr122 <- predict(mR.lo122, data.frame(age = age20to70))

summary(mR.lo122)

plot(mR.lo.pr992 ~ age20to70,
     type = "l", col = "purple", lwd = 3,
     xlab = "Age (years)", ylab = "Fitted Run Time (minutes)")

lines(x = age20to70, y = mR.lo.pr122,
      col = "green", lty = 2, lwd = 3)

### 10.)

We have seen that the 1999 runners were typically older than the 2012 runners. Compare the age distribution of the runners agross all 14 years of the races. Use quantile-quantile plots, boxplots and density curves to make your comparisions.

```{r}
ggplot(cbMen, aes(age, group = year)) +
   geom_histogram(aes(fill = ..count..), bins = 30) +
   facet_wrap(~year)

ggplot(cbMen, aes(age, group = year)) +
   geom_boxplot(aes(fill = as.factor(year))) +
   coord_flip() +
   labs(title = "Age by Race Year", xlab = "Race Year", ylab = "Age")

ggplot(cbMen, aes(year, age)) +
   geom_quantile()
```

### 11.)

Normalize each male runner's time by the fastest time for the runner of the same age. To do this, find the fastest runnerk for each year of age from 20 to 80, tapply() function maybe helpful here. Smooth these times using loess() and find the smoothed fime using predict. Use these smoothed times to normalize each run time.

cbMenSub <- as.data.table(cbMenSub)

fastest_times <- cbMenSub[, .(max = max(runTime)), by = list(year, age)]

normalized_times <- cbMenSub[, .(runTime), by = list(year, age)]
normalized_times <- merge(normalized_times, fastest_times, by = c("year", "age"))
normalized_times[, normTime := runTime/max]

mR.norm <- loess(normTime ~ age, data = normalized_times)

summary(mR.norm)

age_range <- seq( from = 20, to = 80, by = 1 )

smoothed_times <- predict(mR.norm, newdata = age_range)

smooth_results <- data.table(age = age_range, time = smoothed_times)

ggplot(smooth_results, aes(age, time)) +
   geom_point(aes(col = age)) +
   geom_smooth(alpha = .2)
```

### 12.)

Clean the strings in home and menRes to remove all leading and trailing blanks and multiple contiguous blanks. Also, make all letters lower case and remove any punctuation such as '.' or ','.

homeClean2 <- str_remove_all(homeClean, " ")
homeClean2 <- str_remove_all(homeClean2, ",")

head(homeClean2)
```

### 13.)

In section 2.5 we created an id for a runner by pasting together name, year of birth, and state. Consider using the home town instead of the state. How many runners have competed in at least 8 races using this new id?

```{r}
cbMenSub2 <- cbMenSub

cbMenSub2$ID = paste(nameClean, homeClean2, cbMenSub$yob, sep = "_")

races <- tapply(cbMenSub2$year, cbMenSub2$ID, length)

men8 <- cbMenSub2[ cbMenSub2$ID %in% races8, ]

### 14.)

Further refine the set of athletes in the longitudinal analysis by dropping those IDs who have a large jump in time in consecutive races and who did not compete for two or more years in a row. How many unique IDs do you have when you include these additional restrictions?

by_year <- as.data.table(men8)[, .(year, runTime), by = ID]

sans_outliers <- by_year %>%
   group_by(ID) %>%
   mutate(year_gap = year - lag(year, 1), run_diff = runTime - lag(runTime, 1)) %>%
   filter(year_gap <= 2) %>%
   filter(abs(run_diff) < 5) %>%
   distinct(ID)

nrow(sans_outliers)
```
